---
title: "Causal Inference Capstone"
author: "Tina Chen, Cynthia Zhang"
date: "2024-11-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#install.packages("devtools")  
#devtools::install_github("ChihYuChiang/dlsr")
library(dlsr)
library(MASS)
library(dplyr)
library(stringr)
```

Currently, we recreate the simulation process that has been done in the paper https://home.uchicago.edu/ourminsky/Variable_Selection.pdf. 

We first generate a multivariate normal distribution $Z \sim N(0, 1)$ to get the initial variable $Z_{i1} … Z_{ik}$. The correlation between $Z_{i1}$ and $Z_ik$ is $0.7^{|k-1|}$. Then we set the covariate $W_{ik} = Z_{ik}$ for $k ≤ K/2$ and $W_{ik} = 1 (Z_{ik} > 0)$ for $k > K/2$ where $1(Z_{ik} > 0)$ is a indicator function. K is the number of initial variable Z, which is also the number of covariate W. Researching on the causal relationship between dependent variable Y and treatment variable X, we set the truth as $β_1 = 0.5$, coefficients of covariates $W_{i1}$ and $W_{i, K/2+1}$ are 1, and the coefficient of the rest of covariates are 0. So, ideally, the number of variables selected should be 2.


### try double lasso and fit in linear model

```{r}
set.seed(451)
n = 60
K = 40
beta_1 = 0.5

beta_k <- rep(0, K)
beta_k[1] <- 1
beta_k[K / 2 + 1] <- 1

X <- sample(c(rep(1, n / 2), rep(0, n / 2)))

# Generate covariate matrix Z
mu <- rep(0, K)
Sigma <- matrix(0, nrow = K, ncol = K)
for (i in 1:K) {
  for (j in 1:K) {
    Sigma[i, j] <- 0.7^(abs(i - j))  # Correlation = 0.7
  }
}
Z <- mvrnorm(n, mu = mu, Sigma = Sigma)

# Construct W matrix
W <- Z
W[, (K / 2 + 1):K] <- as.numeric(Z[, (K / 2 + 1):K] > 0)

# Generate outcome variable Y
epsilon <- rnorm(n, 0, 1)
Y <- beta_1 * X + W %*% beta_k + epsilon

#  data frame
df <- W %>%
  as.data.frame() %>%
  mutate(X = X, Y = Y)

# Perform double Lasso selection
test <- str_c("V", 1:K)
df_select <- doubleLassoSelect(df = df, outcome = "Y", treatment = "X", test = test)

# Count selected variables
selected_vars_count <- ncol(df_select) - 2 

model_lm <- lm(Y~., data=df_select)
summary(model_lm)


model_full_lm <- lm(Y~., data=df)
summary(model_full_lm)
```

In the first simulation study we have done, we uses 40 covariates (K = 40) in the model. The theoretical model is:
$$Y_i = β_0 + β_1X_i + β_2W_{i1} + β_3W_{i2} + … + β_{41}W_{i,40} + ε_i$$
where $ε_i$ is the error term which is i.i.d with the normal distribution N(0, 1)
Then we applied this model to the simulated data which has 60 observations for 1000 times and the covariates are respectively correlated to $W_{i1}$. After applying double-lasso variable selection, the number of variables that are being selected is 1.17.

```{r}
set.seed(451)
# Define the simulation function
simulation <- function(n , K , beta_1 , beta_k, n_simulations ) {
  
  selected_vars_count <- rep(0, n_simulations)
  
  for (sim in 1:n_simulations) {
    # Generate treatment variable X
    X <- sample(c(rep(1, n / 2), rep(0, n / 2)))
    
    # Generate covariate matrix Z
    mu <- rep(0, K)
    Sigma <- matrix(0, nrow = K, ncol = K)
    for (i in 1:K) {
      for (j in 1:K) {
        Sigma[i, j] <- 0.7^(abs(i - j))  # Correlation = 0.7
      }
    }
    Z <- mvrnorm(n, mu = mu, Sigma = Sigma)
    
    # Construct W matrix
    W <- Z
    W[, (K / 2 + 1):K] <- as.numeric(Z[, (K / 2 + 1):K] > 0)
    
    # Generate outcome variable Y
    epsilon <- rnorm(n, 0, 1)
    Y <- beta_1 * X + W %*% beta_k + epsilon
    
    #  data frame
    df <- W %>%
      as.data.frame() %>%
      mutate(X = X, Y = Y)
    
    # Perform double Lasso selection
    test <- str_c("V", 1:K)
    df_select <- doubleLassoSelect(df = df, outcome = "Y", treatment = "X", test = test)
    
    # Count selected variables
    selected_vars_count[sim] <- ncol(df_select) - 2  # Excluding Y and X
  }
  

  mean(selected_vars_count)
}



result_raw <- simulation(n = 60, K = 40, beta_1 = 0.5, beta_k = beta_k, n_simulations = 1000)



```



In the second simulation study, we uses the same model as before
$$Y_i = β_0 + β_1X_i + β_2W_{i1} + β_3W_{i2} + … + β_{41}W_{i,40} + ε_i$$
The simulated data for this study is basically the same as before, but we additionally set the treatment variable X is associated with one of the covariates $W_{i1}$. After we applied double-lasso, the number of variables that are being selected is 1.24.

```{r}
set.seed(451)
# A function that X is a function of W_1
simulation <- function(n , K , beta_1 , beta_k, n_simulations ) {
  
  selected_vars_count <- rep(0, n_simulations)
  
  for (sim in 1:n_simulations) {
    # Generate treatment variable X
   
    
    # Generate covariate matrix Z
    mu <- rep(0, K)
    Sigma <- matrix(0, nrow = K, ncol = K)
    for (i in 1:K) {
      for (j in 1:K) {
        Sigma[i, j] <- 0.7^(abs(i - j))  # Correlation = 0.7
      }
    }
    Z <- mvrnorm(n, mu = mu, Sigma = Sigma)
    
    # Construct W matrix
    W <- Z
    W[, (K / 2 + 1):K] <- as.numeric(Z[, (K / 2 + 1):K] > 0)
    
    # Generate outcome variable Y
    epsilon <- rnorm(n, 0, 1)
    # X <- sample(c(rep(1, n / 2), rep(0, n / 2)))
    # X is now a function of W_1
    X <- rbinom(n, 1, prob = plogis(W[, 1]))
    Y <- beta_1 * X + W %*% beta_k + epsilon
    
    #  data frame
    df <- W %>%
      as.data.frame() %>%
      mutate(X = X, Y = Y)
    
    # Perform double Lasso selection
    test <- str_c("V", 1:K)
    df_select <- doubleLassoSelect(df = df, outcome = "Y", treatment = "X", test = test)
    
    # Count selected variables
    selected_vars_count[sim] <- ncol(df_select) - 2  # Excluding Y and X
  }
  

  mean(selected_vars_count)
}

# Example usage
K <- 40
beta_k <- rep(0, K)
beta_k[1] <- 1
beta_k[K / 2 + 1] <- 1

result_x_included <- simulation(n = 60, K = 40, beta_1 = 0.5, beta_k = beta_k, n_simulations = 1000)
print(result_x_included)

```



In the third simulation study, we involve in a interaction term between treatment variable X and the covariate $W_{i1}$. The model is:
$$Y_i = β_0 + β_1X_i + β_2W_{i1} + β_3W_{i2} + … + β_{41}W_{i,40} + β_{42}X_i*W_{i1} + ε_i$$
The simulated data is basically the same as the data in the simulation study 1, additionally having the interaction term affect the outcome variable Y. After we applied double-lasso, the number of variables that are being selected is 1.645. 

```{r}

# A function that X is a function of W_1
simulation <- function(n , K , beta_1 , beta_k, n_simulations ) {
  
  selected_vars_count <- rep(0, n_simulations)
  
  for (sim in 1:n_simulations) {
    # Generate treatment variable X
   
    
    # Generate covariate matrix Z
    mu <- rep(0, K)
    Sigma <- matrix(0, nrow = K, ncol = K)
    for (i in 1:K) {
      for (j in 1:K) {
        Sigma[i, j] <- 0.7^(abs(i - j))  # Correlation = 0.7
      }
    }
    Z <- mvrnorm(n, mu = mu, Sigma = Sigma)
    
    # Construct W matrix
    W <- Z
    W[, (K / 2 + 1):K] <- as.numeric(Z[, (K / 2 + 1):K] > 0)
    
    # Generate outcome variable Y
    epsilon <- rnorm(n, 0, 1)

    # X is now a function of W_3
    X <- rbinom(n, 1, prob = plogis(W[, 3]))
    
    # add interaction term: interact with X
    interact_term <- X * W[, 1]
    beta_interact <- 0.2  
    # Y <- beta_1 * X + W %*% beta_k + epsilon
    Y <- 0 + beta_1 * X + W %*% beta_k + beta_interact * interact_term + epsilon
    
    #  data frame
    df <- W %>%
      as.data.frame() %>%
      mutate(X = X, Y = Y, interact = interact_term)
    
    # Perform double Lasso selection
    test <- c(str_c("V", 1:K),"interact")
    df_select <- doubleLassoSelect(df = df, outcome = "Y", treatment = "X", test = test)
    
    # Count selected variables
    selected_vars_count[sim] <- ncol(df_select) - 2  # Excluding Y and X
  }
  

  mean(selected_vars_count)
}

# Example usage
K <- 40
beta_k <- rep(0, K)
beta_k[1] <- 1
beta_k[K / 2 + 1] <- 1

result_interaction <- simulation(n = 60, K = 40, beta_1 = 0.5, beta_k = beta_k, n_simulations = 1000)
print(result_interaction)

```


### Next step
For the future plan, we will add more introduction on double lasso series. We will also add more simulation studies to play around this method.
```{r}
# add non linearity

# Modify beta coefficient, sample size, K
```







